---
title: "John Hopkins Data Science with R"
author: "Mandar Bhosale"
date: "7/15/2019"
output:
  html_document: default
  pdf_document: default
---

# Introduction

This is our first project in R. The learning is acquired through **John Hopkins University Mooc ** *Data Science with R*.

## Contents

We have so far learned how to install R and Rstudio. We took a brief tour of the Rstudiio GUI. Other things covered in the Mooc are:  
- Version Control using Git - [Visit my GitHub Profile](https://github.com/Mandharr)  
- R Markdown  
- Types of Analysis  
- Experimental Design or DOE  
- [R Documentation](www.rdocumentation.org)  

In order to get a brief understanding of the dataset use `summary(dataset name)`

```{r}
print("Hello World")
```

### Descriptive Analysis

Summary of the Data
Ex: Age distribution in US

### Exloratory Analysis

Examine the Data and find relationships that weren't known before
Correlation does not imply causation

Useful for discovering new conneciton
Its just tells that a relation exists, but not the cause

### Inferential Analysis

Small sample of data us used to say something about the population at large. Smaple data plays an important role.
The data should be related.
Ex: Sample data from US population can be used to infer the life expectancy of the pupoulation in US.

### Predictive Analysis

Using current and historic data to predict about the future for the query in hand.
You should be using the right variables and also understand that *correlation does not lead to causation.*

### Causal Analysis

What happens to one variable when we change other.

## Basic Setup

Run RStudio in your Windows/Mac System. 

Type `getwd()` in thr R console to check your current working directory. The working directory is where all data files, R scipts are stored to be used later. 
Using "File" tab -> "..." -> choose folder to open -> More -> Set as working directory.  
Or Got to the folder -> Copy the folder path -> in RStudio console -> type `setwd(pathname)` -> Run.

## R Programming

Data Types :  
- Interger  
- Number  
- Character  
- Logical  

Objects in R  
- Vector  
- Matrix  
- Dataframe  
- List  

Assign variables
Varibales store values. Eg: x <- 1, will store 1 in x.
Values stored in a variables can be printed on the console using the variables by typing `x`. Other way is using `print(x)` function.

`class()` and `mode()` fucntion can help us to understand what type of data is stored in the variables.
```{r}
a <- 1
class(a)
mode(a)
```
```{r}
b <- "R programming"
class(b)
mode(b)
```
Function `c()` helps to concatenate values.
```{r}
x = c(0.5, 0.6)
x = c(1,2,3)
x = c(T,F)
x = c("a","b","c")

```

Vector Funciton `vector()` creates empty vector with specified type.
```{r}
z <- vector("numeric", length = 10)
y <- vector("character", length = 10)
```


Attributes

-  Matrix has dimnames and names  
```{r}
test_mat <- matrix(1:12, nrow = 3, byrow = T)
rownames(test_mat) <- c("1","2","3")
colnames(test_mat) <- c("A","B","C","D")
test_mat
dimnames(test_mat)

```
Naming the Matrix can be done using list, as mentioned later.

- Length of the vector  

Explicit Coercion
which also means: type conversion
`as. * ` is used to convert a object from one to other.  Ex: `x <- 0.6 | as.numeric(x)`. 

### Lists


```{r}
# Multi type vector list
xlist <- list(1,"a", TRUE, 1.1) 
xlist
```
```{r}
# List can have other list, matrix and vector together
ylist <- list(xlist, test_mat)
ylist
```

### Matrices

`dim()` prints the dimensions in the matrix row | columns
`attributes` prints the dimensions aswell the names of the rows and columns

We can create a matrix from a vector. consider vector `man`
```{r}
#dim(man) <- c(2,5) # this converts vector 'man' to a matrix 'man', using the dimmensions provided.
#man
```
Other function in matrix are 
- cbind : Adds columns to a matrix or can be used to add vectors to make a matrix
- rbind : Adds rows 
 
`cb_mat <- cbind(x,y)` or `rb_mat <- rbind(x,y)` 

### Factors

Represents categorical data : Order and Unordered
- Ordered : Male | Female  
- Unordered : Low<Med<High  
Factors have labels

`xfact <- factor(c("yes","no","no","yes","yes))`
```{r}
temp <- c("L","M","H","L","L","M")  # Create a vector
temp_fact   <- factor(temp, levels = c("L","M","H"), ordered = T) # Factor function converts temp variable to factor variable, levels -mentions the order, order - helps give precedence or weightage.
temp_fact
```

### Dealing with missing values

NA or NaN values
NaN are for undefined mathematical operations

using `is.na()` or `is.nan()` helps to find the missing values in the vector
```{r}
xmiss <- c(1,2,NA,3,4,NaN)
is.na(xmiss)
is.nan(xmiss)
```
The output is always logical and gives TRUE when there is a missing value at a location.

### Data Frames

Used to store tabular data. Is like a list and all list have the same length. Column have varied data type.
Attributes are `row.name()` 
We can import data from our system using `read.table()` or `read.csv()`.

To create dataframe : xframe <- data.frame(1:10)

We can create dataframe from vectors.
```{r}
# Create varied vectors 
ID <- c(1:10)
Name <- c("a","b","c","d","e","f","g","h","i","j")
xdata <- data.frame(ID,Name)
xdata
```

This creates a dataframe from the vectors and uses the vector name as column name.  
Other used functions :  
- head(): Gives top 6 rows from a dataframe `head(xdata)`  
- tail(): GIves last 6 rows from the dataframe `tail(xdata)`  
- `xdata$ID` will print only the ID column from the dataframe  
- str(): Will print the structure of the dataframe `str(xdata)`  
- Summary(): Will print the statistical summary of all the numerical values `summary(xdata)`
- subset(): To retrieve data from a dataframe using a condition `subset(xdata, subset = ID>5)`

To add a new column to our dataframe. Simply create a vector or pass vector into a column name not already present in the dataframe.
Ex
```{r}
newcolumn <- c(21:30)
xdata$Marks <- newcolumn
xdata
```

Using `rbind()` we can add a new row to a dataframe.
Ex
```{r}
xdata <- rbind(xdata, c(11,"k",31))
xdata
```

### Name Attributes

R objects like Vector, Matrix, Data.frame has names 

Objects Excample
vector  x <- 1:3 | name(x) <- c("a","b","c")
list    x <- list(a=1, b=2, c=3)
Marix   dimnames(test_mat) <- list(c(provide row names),c(provie col names))


### Reading Data into R

`read.table(), read.csv()` function can be used to read tabular data
`readLines` is used to read lines of a text file.
If we have a R code files that can be used by source() function
load() read binary objects in R

`read.table`is the common.
- file: is the file name or connection  
- header: tells first line is header name  
- sep: indicates how the column are seperated  
- colClasses: indicates the class of each column in the dataset
- nrows: the number of lines in the set
- skip: the number of lines to skip from beginning
- sringsAsFactors: char variabes as factors?

R automatically skips lines begining with #.

Rough Calc says if the dataset is more than ur RAM size, dont use it.

if no comment is present : then comment.char =""

*When reading dataset R goes through each row to set the class. But if we expplicitly mention the class, then things get fast.
```{r}
# initial <- read.table("database.txt", nrows = 100) # get a sample of the dataset
# classes <- sapply(inital, class) # find the class of sample dataser columns
# AllData <- read.table("database.txt", colClasses = classes) # Explicitly mention the class of the columns
```
Knowing the number of Rows also helps

### How to make R runs faster?

- Check system memory  
- other pplication running  
- other users?  
- 

### How  to check the memory required for dataset.

1500000 rows and 120 columns with numeric data
Therefore : 1500000 * 120 * 8
            1440000000 / 2^20
            1.34 GB
Hence we need around twice the memory.

### Textual Formats

dumping and dputing : This is used to read metadata of a dataset.
When in a version control environment this extra information is good for tracking.
"Textual Format data takes a lot of space"

Example
```{r}
ydata <- data.frame(a =  1,b="a") # Create a frame
dput(ydata) # Run dput(). This creates the structure shown below in commment, with the class, row name, col name, etc
# structure(list(a = 1, b = structure(1L, .Label = "a", class = "factor")), class = "data.frame", row.names = c(NA, -1L))
dput(ydata, file = "dputydata.R") # Save the result in a file, to use later
new.ydata <- dget("dputydata.R") # import the file to use
new.ydata
```

dump() : This funciton can be used on multiple R objects as opposed to the dput() shown in example above. and can be retrieved using `source()`

```{r}
x <- "'foo"
y <- data.frame(a=1,b="a")
dump(c("x","y"),file = "data.R")
rm(x,y)
source("data.R") # read from the dump file
x
y
```

## Connections 

- File: Open a file  
- gzfile: compressed gzip file  
- bzfile: bzip2 compressed file  
- url: connect to webpage  

The `file()` structure:  
```{r}
str(file)
```

Description is the name of the file.  
open is to provide the mode we want to open the file. Read, write, Append (rwa) and in binary : rb,wb,ab

con <- file("foo.txt", "r)
data <- read.csv(con)

or data <- read.csv("foo.txt")

the conneciton function can be used for reading specific lines from a file.
con <- gzfile("words.gz")
x<- readLines(con, 10) # This reads the first 10 lines.

`writeLines()` can be used to write the vector to a file, each element on a single line.

### Subsetting

[] : single brackets: subsets objects of the same cass.. therfore , vector gives vector  
[[]]: used with list or dataframe: therfore mulitple classess elements can be subsetted.  
$: When the objects in list or dataframe has *name* we use the dollar sign
```{r}
x <- c(1,2,3,4,5)
x[1]
x[1:4]
x[x > "a"]
u <- x > "a"
x[u]
```
```{r}
y <- list(bar = 0.6)
y$bar
```
```{r}
x <- list(foo = 1:4, bar = 0.6, baz= "hello")
x[c(1,3)] # returns list objects in first and third index : 
# [[c(1,3)]] : Means 1 object in the list, and third element in that object.
```

For Matrix

Excample : x[,1] or x[1,] or x[1,2] ,etc

Removing Missing values in a vector. We can do this using logical values
```{r}
x = c(1,2,NA,4,NA,5)
bad = is.na(x) # stores logical values of values with TRUE in NA element
x[!bad] # prints values where its NOT TRUE for NA
```

To check if two vectors do not have NA in a similar position index use `complete.cases()`. both vector must be same length.
```{r}
x <- c(1,NA,NA,4)
y <- c("a","b",NA,"c")
checkna <- complete.cases(x,y)
checkna
x[checkna]
```

Removing Missing values in dataframe
```{r}
airquality[1:6,]
good <- complete.cases(airquality)
good
airquality[good, ][1:6, ]
```

Complete.Cases is good when we have large datasets, and we want tp sibset out all the missing values

### Vectorized Operations

How this work is, each element of the two vectors are operated upon parallely. 1-1,2-2, etc. If the one vector has les elements then then vector with
less elemenent will be used in the alternate manner, 1st element , then 2nd , then agin 1st for the calculations

x<-1:4 | y<- 6:9 Therefore, 1+6,2+7, etc. This also works in matrix.
If the vectors are of different lengths, R 'recycles' the shorter vector until it is the same length as the longer vector.
when you ask R to compute z * 2 + 100,what it really computes is this: z * c(2, 2, 2) + c(100, 100, 100).

## Installing SWIRL

### Statistics With Interactive R Learning  
`install.packages("swirl")`
`library(swirl)`

### Swirl Learning

R Workspace
- abs() : to get the absolute value of number  
- sqrt() : square root.  
- getwd(): get present woking directory  
- ls() : list all objects in the current work environment.  
- dir(): lists all files in the current working directory.| list.files().  
- args(): to check the arguments a function can take.  
- dir.create(): create a new directory.`dir.create("testdir)` in pwd.  
- setwd(): set a dir as working. `setwd("testdir")`.  
- file.create(): create a file in pwd. `file.create("mytest.R")`  
- file.exists(): check if a file is present in a directory. `file.exists("mytest.R")`  
- file.info().  
- file.rename().  
- file.copy().  
- file.path("")
- file.path('folder1'): create platfrom independent file path.  
# dir.create(file.path("testdir2","testdir3"), recursive = TRUE) #Create a folder in folder as directory  

Sequence of Numbers
- 1:20.integer  
- pi:10.real numbers  
- 15:1 descending.  
- seq(): seq(1,20)  
- seq(0,10,by=0.5)  
- seq(5,10,length=30): this gives 30 numbers between 5-10.  
- length(): to find the length of a vector.
# 1:length(my_seq).  gives vector of length 1-myseq length.
- seq(along.with = myseq)
- seq_along(myseq)
- rep(0, times=40) : replicate 0, 40 times  
- rep(c(0,1,2), times = 10)  
- rep(c(0,1,2), each=10). Each element 10 times  

Vectors

Atomic Vectors or Multi Data type, called List.
Logical Vectors
  num_vect <- c(0.5, 55, -10, 6)
  tf <- num_vect < 1
  tf

here tf has vector values telling if the condition of  < 1 was satisfied by each each element of num_vect 

Logical operator : < > =< => == != A|B A&B 

Character vector:
- paste(): merges the values in a vector in to 1. 
# paste(my_char, collapse = " "). collapse tell what needs to be there between them when forming a string.  
- concat: c(mychar, "new word")  
- paste(1:3, c("X","Y","Z"), sep = "").  

Missing Values  
- NA is 'not available'. Anything multiplied by NA is NA.
- `my_data <- sample(c(y,z), 100)`:  running `my_na <- is.na(my_data)` will give where all the NA are located in the my_data.

All location with TRUE means, that element is NA.
To count number of missing values . `sum(my_na)` since TRUE is asigned to NA, and TRUE is 1, sum will count all the NAs.

NaN stands for Not a Number. 0/0 or Inf - Inf.


Creating draws from standard disribution
`rnorm(1000)` returns 1000 draws in random.

Getting a random sample size from a vector
`y <- rnorm(1000)`
`z <- rep(NA, 1000)`
`my_data <- sample(c(y,z), 100)` # Grabs mix values from y and z totaling to 100.


Subsetting Vectors

Extracting by indexing with logical vectors.
x[is.na(x)] : since is.na gives TRUE for Na, X[] will print all NA for every TRUE received

To get values that are not NA, we can negate the function: y <- x[!is.na(x)].  

Its is important to remove all NA first and then find positive values, As logical operator will bring in NA as a output. 
```{r}
x[!is.na(x)]
y <- x[!is.na(x)]
# # y
#  [1] -1.26763232  0.11205691 -1.42334483  0.50911397 -1.45539683 -0.58333960 -1.60977351 -0.28791459 -0.77927997  2.36320449 -0.63959542  0.29379196 -1.53424020
# [14]  0.46931339 -2.27577209  1.47792103  0.03453443 -0.69009558 -0.14828570 -0.86818327

# y[y>0]
# [1] 0.11205691 0.50911397 2.36320449 0.29379196 0.46931339 1.47792103 0.03453443

# x[x>0]
#  [1] 0.11205691         NA         NA         NA 0.50911397         NA         NA         NA         NA 2.36320449         NA         NA 0.29379196         NA
# [15] 0.46931339         NA         NA         NA         NA 1.47792103 0.03453443         NA         NA         NA         NA         NA         NA
```
# x[!is.na(x) & x > 0] in one line code.

x[c(-2, -10)] or x[-c(2,10)]

names(vect) <- c("foo", "bar","norf") # This gives names to the vectors

To check if two vectors are same : idential(vect1, vect2)  
$ ***********************************************************$

Matrices and Data Frames

dim(my_vector)
dim(my_vector) <- c(4,5) helps to set the dimensions of the vector.
attributes(my_vector) returns the attributes of the vector inclunding the dim.

Create a matrix: my_matrix2 <- matrix(1:20, nrow = 4, ncol = 5)  
Create a vector: patients <- c("Bill", "Gina", "Kelly", "Sean")
Add vector to matrix as a column: cbind(patients, my_matrix)

To give column names : cnames <- c("patient", "age","weight", "bp", "rating","test")
`colnames(my_data) <- cnames` This adds names to every column.


## Control Structures 

- if, else: testing a condition  
- for: execute loop fixed number of times  
- while: loop while condition is true  
- repeat: infinite loop  
- break: break a loop  
- next: skip an iteration of a loop.  
- return: exit a function  


### IF - ELSE

if()
{
}
else

```{r}
if(x>3) {
  y <- 10
} else {
  y <- 0
  
}
```
```{r}
y <- if(x > 3){
  10
} else {
  0
}
```

Just IF :

if(){}  # this works fine.

### FOR 

It is the most common type of loop. below are some of the ways we can do the same job
```{r}
for(i in 1:10) {
  print(i)
}
```
```{r}
x <- c("a","b","c","d")

for(i in 1:4) {
  print(x[i])
}

for(i in seq_along(x)) {
  print(x[i])
}

for(letter in x){
  print(letter)
}

for (i in 1:4) print(x[i]) 
```

Nested For Loop

```{r}
x <- matrix(1:6,2,3)

for(i in seq_len(nrow(x))) {
  for(j in seq_len(ncol(x))) {
    print(x[i,j])
  }
}
```

### WHIle Loop

```{r}
count <- 0
while(count < 10) {
  print(count)
  count <- count +1 
}
```

```{r}
z <- 5

while(z >= 3 && z <= 10) {
  print(z)
  coin <- rbinom(1,1,0.5) # random number generation
  
  if(coin ==1) {
    z<- z+1
  } else {
    z <- z -1
  }
}
```

### Repeat, Next & Break

```{r}
for (i in 1:100) {
  if(i <= 20){
    next # skips number till 20 
    
  }
  print(i)
}
```
Returns tell to exit the function and return the last value

# Lets Write Function 

Function are used a lot in data or statistical analysis .
Its always good to write function in a text file for reproducibility.

```{r}
add2 <- function(x,y) { # function to create, accepts two values and returns the result after computing
  x + y # tells what to do with the values accepted
} 
  
```
```{r}
x <- 1:20
aboveten <- function(x) { # takes a vector from user and send back values greater than 10
  use <- x > 10
  x[use]
}

above <- function(x, n) { # similar to aboveten function, but this also accepts the condition for subsetting , it can also be default :n=10
  use <- x > n
  x[use]
}
```

Accept matrix nd find mean

```{r}
columnmean <- function(y) {
  nc <- ncol(y) # find the number of column in the matrix
  means <- numeric(nc) # making sure the number are numeric for looping
  
  for(i in 1:nc) { # create a for-loop of i from 1 to columns in matrix
    means[i] <- mean(y[,i]) # means[i] :create a dynamic vector | mean(y[,i]) mean of the [i] column 
  }
  means # return mean
}

columnmean(airquality) # passing the airquality dataframe
```

They Belong to the class : Function
We can nest funcion and be passed in other function.

Named arguments have default values
formals  : returns a list of all fornal arguuments
Argument matching is done positionally or by name.

Which means :
```{r}
mydata <- 1:20
sd(mydata)
sd(x=mydata)
sd(x= mydata, na.rm=FALSE)
sd(na.rm = FALSE, x=mydata)
# they all mean the same .
```
USing args() can tell you what arguments are necesaary and what are there locations.

We can have function with argument value = NULL

### Lazy Evaluation

Arguments to functions are evaluated only as needed.

```{r}
func <- function(a,b){ # Even if we need 2 arguments, R can function with 1. 
  a^2                  # Since R does nt use value of B, there is no error
}

func(2)
```
BUt, if R needs B and B is not passed by user this will produce error.

```{r}
func <- function(a,b) {
  print(a)  # A will be executed
  print(b) # Since R is using B, console will gibe error of missing value
}

func(45)
```

The ... Arguments 

We can use argument list from other function, in our custom function.

```{r}
myplot <- function (x,y,type ="l", ...) { # We are customizing plot function and setting default type to "L", and using every other argument
  plot(x,y,type =type, ...)
}
```

Scoping Rules of R

When we are working in R, and if there are functions with same name, R takes the ordered way of using the funciton. Which starts from Global Environment.
`search()` function can be used to know the order.


# Lexical Scoping 

```{r}
f <- function(x,y) {  # x and Y are formal arguments.
  x^2 + y /z          # Here z is not defined but used. Z is a free variable
}

```

Scoping rule of R determines what value needs to be assigned to free variable. They are not local variables 
if the value is not found in environment, it moves to parent environemt unitl reaches the global environment. 
And Still contniues to reach the empty environment. And if still not found " ERROR OCCURS"

```{r}
make.power <- function(n) {
  pow <- function(x) {
    x^n
  }
  pow
}
cube <- make.power(3) #We can make a function act differently based on value we provide
cube(3) 

cube <- make.power(2)
cube(3)
```

How to know what the function environemnt ?

`ls(environment(cube))` returns "n" and "pow"
`get("n",environment(cube))` returns value of n


Optimization

Write a Constructor


