---
title: "John Hopkins Data Science with R"
author: "Mandar Bhosale"
date: "7/15/2019"
output:
  html_document: default
  pdf_document: default
---

# Introduction

This is our first project in R. The learning is acquired through **John Hopkins University Mooc ** *Data Science with R*.

## Contents

We have so far learned how to install R and Rstudio. We took a brief tour of the Rstudiio GUI. Other things covered in the Mooc are:  
- Version Control using Git - [Visit my GitHub Profile](https://github.com/Mandharr)  
- R Markdown  
- Types of Analysis  
- Experimental Design or DOE  
- [R Documentation](www.rdocumentation.org)  

In order to get a brief understanding of the dataset use `summary(dataset name)`

```{r}
print("Hello World")
```

### Descriptive Analysis

Summary of the Data
Ex: Age distribution in US

### Exloratory Analysis

Examine the Data and find relationships that weren't known before
Correlation does not imply causation

Useful for discovering new conneciton
Its just tells that a relation exists, but not the cause

### Inferential Analysis

Small sample of data us used to say something about the population at large. Smaple data plays an important role.
The data should be related.
Ex: Sample data from US population can be used to infer the life expectancy of the pupoulation in US.

### Predictive Analysis

Using current and historic data to predict about the future for the query in hand.
You should be using the right variables and also understand that *correlation does not lead to causation.*

### Causal Analysis

What happens to one variable when we change other.

## Basic Setup

Run RStudio in your Windows/Mac System. 

Type `getwd()` in thr R console to check your current working directory. The working directory is where all data files, R scipts are stored to be used later. 
Using "File" tab -> "..." -> choose folder to open -> More -> Set as working directory.  
Or Got to the folder -> Copy the folder path -> in RStudio console -> type `setwd(pathname)` -> Run.

## R Programming

Data Types :  
- Interger  
- Number  
- Character  
- Logical  

Objects in R  
- Vector  
- Matrix  
- Dataframe  
- List  

Assign variables
Varibales store values. Eg: x <- 1, will store 1 in x.
Values stored in a variables can be printed on the console using the variables by typing `x`. Other way is using `print(x)` function.

`class()` and `mode()` fucntion can help us to understand what type of data is stored in the variables.
```{r}
a <- 1
class(a)
mode(a)
```
```{r}
b <- "R programming"
class(b)
mode(b)
```
Function `c()` helps to concatenate values.
```{r}
x = c(0.5, 0.6)
x = c(1,2,3)
x = c(T,F)
x = c("a","b","c")

```

Vector Funciton `vector()` creates empty vector with specified type.
```{r}
z <- vector("numeric", length = 10)
y <- vector("character", length = 10)
```


Attributes

-  Matrix has dimnames and names  
```{r}
test_mat <- matrix(1:12, nrow = 3, byrow = T)
rownames(test_mat) <- c("1","2","3")
colnames(test_mat) <- c("A","B","C","D")
test_mat
dimnames(test_mat)

```

- Length of the vector  

Explicit Coercion
which also means: type conversion
`as. * ` is used to convert a object from one to other.  Ex: `x <- 0.6 | as.numeric(x)`. 

### Lists


```{r}
# Multi type vector list
xlist <- list(1,"a", TRUE, 1.1) 
xlist
```
```{r}
# List can have other list, matrix and vector together
ylist <- list(xlist, test_mat)
ylist
```

### Matrices

`dim()` prints the dimensions in the matrix row | columns
`attributes` prints the dimensions aswell the names of the rows and columns

We can create a matrix from a vector. consider vector `man`
```{r}
#dim(man) <- c(2,5) # this converts vector 'man' to a matrix 'man', using the dimmensions provided.
#man
```
Other function in matrix are 
- cbind : Adds columns to a matrix or can be used to add vectors to make a matrix
- rbind : Adds rows 
 
`cb_mat <- cbind(x,y)` or `rb_mat <- rbind(x,y)` 

### Factors

Represents categorical data : Order and Unordered
- Ordered : Male | Female  
- Unordered : Low<Med<High  
Factors have labels

`xfact <- factor(c("yes","no","no","yes","yes))`
```{r}
temp <- c("L","M","H","L","L","M")  # Create a vector
temp_fact   <- factor(temp, levels = c("L","M","H"), ordered = T) # Factor function converts temp variable to factor variable, levels -mentions the order, order - helps give precedence or weightage.
temp_fact
```

### Dealing with missing values

